State of the Practice of Lattice Boltzmann Method Software Development

### 2021-06-16 StateOfPractice/Peter-Notes/Report/Report.tex
applied first set of feedback, specifically Methodology section

The LBM software packages are assessed for a given set of quality attributes, using quantitative and qualitative data, in order to answer the research questions found in Section \ref{researchquestions}. The qualities are defined in Section \ref{softwarequalities}. Qualitative data is gathered from interviews with software package developers. The interview questions are listed in Section \ref{interviewquestions}. Quantitative data is measured using the quality metrics listed in Section \ref{measurementtemplate2}. In this context the purposes of the report are as follows:

@@

In the course of this assessment we updated the methodology that was used during previous assessments. Details of the new methodology are in \citep{methodology2020}. The previous set of research questions were critically assessed and modified. The updated questions are listed in Section \ref{researchquestions}. In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, and added more measures, including collecting empirical data and interviewing developers. We have also added a domain analysis to better characterize the functionality provided by the software, and have leveraged the expertise of a domain expert. As in past assessments, the collected data was combined to rank the software using the Analytic Hierarchy Process (AHP). The domain expert was consulted to verify the ordering.

@@

This report analyzes a filtered set of LBM software packages along quantitative and qualitative measures. Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality; they may not represent the true quality of the software. Surface measurements are taken as they allow us to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities among the set of software packages. The measured software packages are open source, and it is recommended to keep this in mind when considering the recommendations of this report. Some recommendations may not apply to closed-source software. Practices surrounding close source software development may differ.

@@

In this project we set out to answer the research questions found in Section \ref{researchquestions} for the LBM SCS domain. The process involved systematically measuring and analyzing members of this software family along the quality attributes found in Section \ref{softwarequalities}. This included gathering quantitative and qualitative measurements. A goal of the project was to produce a hierarchical quality assessment for LBM software packages.

We collected quantitative data using the measures found in the measurement template of Section \ref{measurementtemplate}. Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed. Most of the data was gathered by manually observing the software, its source code, and its artifacts, while some was gathered using the tools discussed in Section \ref{empiricalmeasures}.

We also collected qualitative data by interviewing the software package developers and asking them the questions found in Section \ref{interviewquestions}. Furthermore, we solicited the assistance of domain experts to better assess each software package by leveraging their experience to assess the functional and non-functional requirements for the software domain.

@@

\begin{enumerate}
	\item List candidate software packages for the domain. This is discussed in Section \ref{identifysoftware}.
	\item Filter the software package list. This is discussed in Section \ref{filtersoftware}.
	\item Gather the source code and documentation for each software package.
	\item Collect empirical measures. This is discussed in Section \ref{empiricalmeasures}.
	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in Section \ref{measurementtemplate2}.
	\item Survey the developers. The developers of each software package in the filtered software list were contacted for voluntary interviews. Four interviews were conducted in total. The interview questions can be found in Section \ref{interviewquestions}.
	\item Use AHP to rank the software packages. This is discussed in Section \ref{AHP}.
	\item Analyze the results and answer the research questions. The answers can be found in section \ref{answersquestions}.
\end{enumerate}

These steps are further detailed in the Methodology for Assessing the State of the Practice for Domain X document found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{here}.

@@

Software quality attributes facilitate the measurement and comparison of software packages in this state of the practice assessment. In this report we adopt software quality definitions from various researchers and subject matter expert entities. Some of the definitions are from \cite{Smithetal2020}. The quality measurement results are found in Section \ref{qualityresults} of this report. Section \ref{qualityrecommentations} lists recommendations to address software qualities in LBM software packages. The following are the software quality definitions used in this state of the practice exercise, along with comments regarding their quantitative and qualitative measurement.

@@

 A software program is correct if it behaves according to its stated
specifications \citep{GhezziEtAl2003}. This requires that the specification is available. Software is less likely to have a formal specification if it is not developed by seasoned or professional software developers. Since some software does not have a specification available, the correctness of software cannot always be verified. Despite an absent specification, the correctness of the output of scientific computing software can sometimes be manually verified by applying domain knowledge. A good measure of correctness correlates with the availability of a requirements specification and reference to domain theory, as well as the explicit use of tools or techniques for building confidence of correctness, such as documentation generators and software analysis tools.

@@

Reliability is measured by the probability of failure-free operation of a computer program in a specified environment for a specified time, i.e. the average time interval between two failures also known as the mean time to failure (MTTF) \citep{GhezziEtAl2003} \citep{musa1987software}. Reliability is thus positively correlated with the absence of errors during installation and use. Recoverability from errors also improves reliability. In turn the presence of descriptive error messages and logs increases the ability to recover the system.

@@

Understandability is measured by the capability of the software package to enable the user to understand its suitability and function \citep{ISO9126}.
It is an artifact-dependent quality. Understandability is different for the user-interface, source code, documentation. In this state of the practice analysis understandability focuses on the source code. It is measured by the consistency of a formatting style, the extend of modularization, the explicit identification of coding standards, the presence of meaningful identifiers, and clarity of comments.

@@

Software achieves reproducibility if another developer can take the requirements documentation and re-obtain the same software artifacts \citep{BenureauAndRougier2017}. In this context, reproducibility refers to the similarity between artifacts. Reproducibility can also refer to the output of the software, irrespective of the artifacts, where the scientific results are compared between software implementations, or between software implementations and manually calculated results. This state of the practice assessment did not quantitatively measure reproducibility. Developers were asked in interviews if they have any concern that their computational results won’t be reproducible in the future, and if they have taken any steps to ensure reproducibility.

@@

Unambiguity refers to the extent to which two readers have similar interpretations when reading software artifacts. In other words, artifacts are unambiguous if, and only if, they only have one interpretation \citep{IEEE1998}. This state of the practice assessment did not quantitatively measure unambiguity. Developers were asked in interviews if they think that the current documentation can clearly convey all necessary knowledge to the users, and how they achieved this or what improvements are needed ot achieve it.

@@

The initial list had 46 packages, including a few packages that were later found to not have publicly available source code, or to be in an incomplete state of development.

@@

In order to reduce the number of members in the candidate software list, the following filters were applied. The filters were applied in the priority order listed.

@@

\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\hline
		Name & Released & Updated\\
		\hline
		DL\_MESO (LBE) & unclear & 2020 Mar\\
		waLBerla & 2008 Aug & 2020 Jul\\
		Sailfish & 2012 Nov & 2019 Jun\\
		Palabos & unclear & 2020 Jul\\
		OpenLB & 2007 Jul & 2019 Oct\\
		MechSys   & 2008 Jun    & 2020 Jul\\
		LUMA&   2016 Nov   & 2020 Feb\\
		Ludwig& 2018 Aug & 2020 Jul\\
		lettuce    & 2019 May & 2020 Jul\\
		pyLBM & 2015 Jun&   2020 Jun\\
		lbmpy& unknown  & 2020 Jun  \\
		TCLB & 2013 Jun  & 2020 Apr\\
		ESPResSo & 2010 Nov & 2020 Jun\\
		ESPResSo++ & 2011 Feb & 2020 Apr\\
		\hline
	\end{tabular}
	\captionof{table}{Alive Software Packages}\label{alivepackages}
\end{center}

While the initial list had 46 packages, filtering by scope, usage, and age decreased the size of the list to 23 packages. Many of the 23 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, or the project was out of scope or not up to a standard that would support incorporating them into this study. Of the remaining 23 packages that were studied, some were kept on the list despite being marked as dead due to their prevalence on authoritative lists on LBM software and due to their surface excellence, specifically the considerable time that was put into these projects.  
The final list of software packages that were analyzed in this project can be found in the following two tables. Table \ref{alivepackages} lists packages that fell into the `alive' category as of mid 2020, and Table \ref{deadpackages} lists packages that were `dead' at that time. \\

\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\hline
		Name & Released & Updated\\
		\hline
		SunlightLB & 2005 Sep & 2012 Nov\\
		MP-LABS & 2008 Jun & 2014 Oct\\
		LIMBES & 2010 Nov & 2014 Dec\\
		LB3D & unclear & 2012 Mar\\
		LB3D-Prime & 2005 & 2011 Oct\\
		LB2D-Prime & 2005 & 2012 Apr\\
		LatBo.jl & 2014 Aug & 2017 Feb\\
		HemeLB & 2007 Jun & 2018 Aug\\
		laboetie & 2014 Nov & 2018 Aug\\
		\hline
	\end{tabular}
	\captionof{table}{Dead Software Packages}\label{deadpackages}
\end{center}

@@

The quality measurements in this assessment rely on the gathering and analyzing of raw and processed empirical data related to the research questions listed in Section \ref{researchquestions}.

All of the quality measurements that are part of the AHP analysis are in this category, while qualitative data gathered during interviews with developers is not. This part of the assessment focuses on data that is reasonably easy to collect. Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using the freeware tools discussed below. The data that was gathered is listed in the measurement template found in Section \ref{measurementtemplate2}. Some of the data required processing other data within the template, including the status of the software package which relied on the last commit date, the percentage of issues that are closed which relied on the number of open and closed issues, and the percentage of code that is comments which relied on the number of total lines and comment lines in text-based files. The complete measurement template data was then analyzed using the grading template found in Section \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.

@@

The Analytical Hierarchy Process (AHP) is a decision-making technique that is used to compare multiple options by multiple criteria. In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in Section \ref{measurementtemplate2} using the grading template found in Section \ref{gradingtemplate2}. AHP performs a pairwise analysis using a matrix and generates an overall score as well as individual quality scores for each software package. \cite{SmithEtAl2016} shows how AHP is applied to ranking software based on quality measures.

This project used a tool for conducting this process. The tool includes a sensitivity analysis that was used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores. The README file of the tool, which includes requirements and usage information, is found \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{here}.

@@

>>>>

The LBM software packages are assessed in order to answer the research questions found in the Appendix in \ref{researchquestions}. They are assessed for a given set of quality attributes using quantitative and qualitative data. The qualities are defined in Section \ref{softwarequalities}. Qualitative data is gathered from interviews with software package developers. The interview questions are listed in the Appendix in \ref{interviewquestions}. Quantitative data is measured using the quality metrics listed in the Appendix in \ref{measurementtemplate2}. In this context the purposes of the report are as follows:

@@

In the course of this assessment we updated the methodology that was used during previous assessments. Details of the new methodology are in \citep{methodology2020}. The previous set of research questions were critically assessed and modified. The updated questions are listed in the Appendix in \ref{researchquestions}. In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, and added more measures, including collecting empirical data and interviewing developers. We have also added a domain analysis to better characterize the functionality provided by the software, and have leveraged the expertise of a domain expert. As in past assessments, the collected data was combined to rank the software using the Analytic Hierarchy Process (AHP). The domain expert was consulted to verify the ordering.

@@

This report analyzes a filtered set of LBM software packages along quantitative and qualitative measures. Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality; they may not represent the true quality of the software. Surface measurements are taken as they allow us to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities among the set of software packages. Not all qualities are quantitatively measured. For example, performance is not measured by running the code. A surface investigation of the documentation of each software package for performance information is conducted instead. The measured software packages are open source, and it is recommended to keep this in mind when considering the recommendations of this report. Some recommendations may not apply to closed-source software. Practices surrounding close source software development may differ.

@@

In this project we set out to answer the research questions found in the Appendix in \ref{researchquestions} for the LBM SCS domain. The process involved systematically measuring and analyzing members of this software family along the quality attributes found in Section \ref{softwarequalities}. This included gathering quantitative and qualitative measurements. A goal of the project was to produce a quality assessment for LBM software packages.

The methodology used in this LBM software assessment is a general state of the practice methodology that can be applied to any SCS domain. It was developed as an update to previous state of the practice exercises. This general methodology is linked in the next subsection.

We collected quantitative data using the measures found in the measurement template of Section \ref{measurementtemplate}. Some of this was empirical software engineering related data, such as the number of files, number of lines of code, percentage of issues that are closed, etc. Most of the data was gathered by manually investigating the software, its source code, and its artifacts, while some was gathered using the empirical measurement tools discussed in Section \ref{empiricalmeasures}.

We also collected qualitative data by interviewing the software package developers and asking them the questions found in the Appendix in \ref{interviewquestions}. Furthermore, we solicited the assistance of domain experts to better assess each software package by leveraging their experience to assess the functional and non-functional requirements for the software domain.

@@

\begin{enumerate}
	\item List candidate software packages for the domain. This is discussed in Section \ref{identifysoftware}.
	\item Filter the software package list. This is discussed in Section \ref{filtersoftware}.
	\item Gather the source code and documentation for each software package.
	\item Collect empirical measures. This is discussed in Section \ref{empiricalmeasures}.
	\item Measure using the measurement template. This is discussed in Section \ref{empiricalmeasures}. The measurement template can be found in the Appendix in \ref{measurementtemplate2}.
	\item Survey the developers. The developers of each software package in the filtered software list were contacted for voluntary interviews. The interview questions can be found in the Appendix in \ref{interviewquestions}.
	\item Use AHP to rank the software packages. This is discussed in Section \ref{AHP}.
	\item Analyze the results and answer the research questions. The answers can be found in Section \ref{answersquestions}.
\end{enumerate}

These steps are further detailed in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document.

@@

Software quality attributes facilitate the measurement and comparison of software packages in this state of the practice assessment. We adopt software quality definitions from various researchers and subject matter expert entities. Some of the definitions are from \cite{Smithetal2020}. The quality measurement results are found in Section \ref{qualityresults} of this report. Section \ref{qualityrecommentations} lists recommendations to address software qualities in LBM software packages. The following are the software quality definitions used in this state of the practice exercise, along with comments regarding their quantitative and qualitative measurement.

@@

 A software program is correct if it behaves according to its stated
specifications \citep{GhezziEtAl2003}. This requires that the specification is available. Software is unlikely to have a formal specification if it is not developed by seasoned or professional software developers. Since some software does not have a specification available, the correctness of software cannot always be verified. Despite an absent specification, the correctness of the output of scientific computing software can sometimes be manually verified by applying domain knowledge. A good measure of correctness correlates with the availability of a requirements specification and reference to domain theory, as well as the explicit use of tools or techniques for building confidence of correctness, such as documentation generators and software analysis tools.

@@

Reliability is measured by the probability of failure-free operation of a computer program in a specified environment for a specified time, i.e. the average time interval between two failures also known as the mean time to failure (MTTF) \citep{GhezziEtAl2003} \citep{musa1987software}. Reliability is thus positively correlated with the absence of errors during installation and use. Recoverability from errors also improves reliability.

@@

Understandability is measured by the capability of the software package to enable the user to understand its suitability and function \citep{ISO9126}.
It is an artifact-dependent quality. Understandability is different for the user-interface, source code, and documentation. In this state of the practice analysis, understandability focuses on the source code. It is measured by the consistency of a formatting style, the extend of modularization, the explicit identification of coding standards, the presence of meaningful identifiers, and clarity of comments.

@@

Software achieves reproducibility if another developer can take the requirements documentation and re-obtain the same software artifacts \citep{BenureauAndRougier2017}. This includes the output of the software, where the scientific results are compared between software implementations, or between software implementations and manually calculated results. This state of the practice assessment did not quantitatively measure reproducibility. Developers were asked in interviews if they have any concern that their computational results won't be reproducible in the future, and if they have taken any steps to ensure reproducibility.

@@

Unambiguity refers to the extent to which two readers have similar interpretations when reading software artifacts. In other words, artifacts are unambiguous if, and only if, they only have one interpretation \citep{IEEE1998}. This state of the practice assessment did not quantitatively measure unambiguity. Developers were asked in interviews if they think that the current documentation can clearly convey all necessary knowledge to the users, and how they achieved this or what improvements are needed to achieve it.

@@

The initial list had 45 packages, including a few packages that were later found to not have publicly available source code, or to be in an incomplete state of development.

@@

To reduce the number of members in the candidate software list to a manageable size, the following filters were applied. The filters were applied in the priority order listed.

@@

\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\hline
		Name & Released & Updated\\
		\hline
		DL\_MESO (LBE) & unclear & 2020 Mar\\
		ESPResSo & 2010 Nov & 2020 Jun\\
		ESPResSo++ & 2011 Feb & 2020 Apr\\
		lbmpy& unknown  & 2020 Jun  \\
		lettuce & 2019 May & 2020 Jul\\
		Ludwig& 2018 Aug & 2020 Jul\\
		LUMA& 2016 Nov   & 2020 Feb\\
		MechSys & 2008 Jun    & 2020 Jul\\
		OpenLB & 2007 Jul & 2019 Oct\\
		Palabos & unclear & 2020 Jul\\
		pyLBM & 2015 Jun&   2020 Jun\\
		Sailfish & 2012 Nov & 2019 Jun\\
		TCLB & 2013 Jun  & 2020 Apr\\
		waLBerla & 2008 Aug & 2020 Jul\\
		\hline
	\end{tabular}
	\captionof{table}{Alive Software Packages}\label{alivepackages}
\end{center}

While the initial list had 45 packages, filtering by scope, usage, and age decreased the size of the list to 23 packages. Many of the 22 packages that were removed could not be tested as there was no installation guide, they were incomplete, source code was not publicly available, a license was needed, or the project was out of scope or not up to a standard that would support incorporating them into this study. These eliminated software packages are listed in the Appendix in Section \ref{eliminatedpackagessection}. Of the remaining 23 packages that were studied, some were kept on the list despite being marked as dead due to their prevalence on authoritative lists on LBM software and due to their surface excellence, specifically the considerable time that was put into these projects.

The final list of software packages that were analyzed in this project can be found in the following two tables. Table \ref{alivepackages} lists packages that fell into the `alive' category as of mid 2020, and Table \ref{deadpackages} lists packages that were `dead' at that time. \\

\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\hline
		Name & Released & Updated\\
		\hline
		HemeLB & 2007 Jun & 2018 Aug\\
		laboetie & 2014 Nov & 2018 Aug\\		
		LatBo.jl & 2014 Aug & 2017 Feb\\
		LIMBES & 2010 Nov & 2014 Dec\\
		LB3D & unclear & 2012 Mar\\
		LB2D-Prime & 2005 & 2012 Apr\\
		LB3D-Prime & 2005 & 2011 Oct\\
		MP-LABS & 2008 Jun & 2014 Oct\\
		SunlightLB & 2005 Sep & 2012 Nov\\
		\hline
	\end{tabular}
	\captionof{table}{Dead Software Packages}\label{deadpackages}
\end{center}

@@

The quality measurements in this assessment rely on the gathering and analyzing of raw and processed empirical data related to the research questions listed in the Appendix in \ref{researchquestions}.

All of the quality measurements that are part of the AHP analysis are empirical measurements. Qualitative data gathered during interviews with developers is not part of the AHP analysis. This part of the assessment focuses on data that is reasonably easy to collect. Much of the data was gathered by manually reviewing the artifacts of each software package, while some was gathered using the freeware tools discussed below. The data that was gathered is listed in the measurement template found in the Appendix in \ref{measurementtemplate2}. Some of the data required processing other data within the template, including the status of the software package, which relied on the last commit date; the percentage of issues that are closed, which relied on the number of open and closed issues; and the percentage of code that is comments, which relied on the number of total lines and comment lines in text-based files. The complete measurement template data was then analyzed using the grading template found in the Appendix in \ref{gradingtemplate2}, the output of which was analyzed using the AHP described in Section \ref{AHP}.

@@

The Analytical Hierarchy Process (AHP) is a decision-making technique that is used to compare multiple options by multiple criteria. In our work AHP was used for comparing and ranking the LBM software packages using the overall impression quality scores that were gathered in the measurement template found in the Appendix in \ref{measurementtemplate2} using the grading template found in the Appendix in \ref{gradingtemplate2}. AHP performs a pairwise analysis using a matrix and generates an overall score as well as individual quality scores for each software package. \cite{SmithEtAl2016} shows how AHP is applied to ranking software based on quality measures.

This project used a tool for conducting this process. The tool includes a sensitivity analysis that was used to ensure that the software package rankings are appropriate with respect to the uncertainty of the quality scores. The \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README} file of the tool includes requirements and usage information.

@@

\subsection{Eliminated Software Packages}\label{eliminatedpackagessection}

The table below lists the software packages that were filtered out of the candidate software list.

\begin{center}
	\begin{tabular}{ p{4cm}p{4cm}}
		\hline
		Name & Reason(s) For Removal\\
		\hline
		ch4-project&Scope \& Usage\\
		CUDA-LBM-simulator&Scope \& Age\\
		elbe&Scope \& Usage\\
		eLBM&Usage\\
		firesim&Scope \& Usage \& Age\\
		fvLBM&Scope \& Usage \& Age\\
		JFlowSim&Scope \& Usage \& Age\\
		LatticeBoltzmann&Scope \& Usage\\
		LBM&Scope \& Usage \& Age\\
		LBM-Cplusplus&Scope \& Usage\\
		LBM-EP&Usage\\
		LBM\_MATLAB&Scope \& Usage \& Age\\
		LBSim&Scope \& Usage \& Age\\
		listLBM&Scope \& Usage \& Age\\
		loliverhennigh&Scope \& Usage \& Age\\
		openLBMflow&Scope \& Usage \& Age\\
		ParallelLbmCranfield&Usage \& Age\\
		PowerFLOW&Usage\\
		ProLB&Usage\\
		Taxila-LBM&Usage \& Age\\
		turbulent\_lbm\_multigpu&Usage \& Age\\
		wlb&Scope \& Usage \& Age\\
		\hline
	\end{tabular}
	\captionof{table}{Eliminated Software Packages}\label{eliminatedpackages}
\end{center}
